---
title: "Analysis of California Schools Test Data"
author: "Zaman (adopted from R-bloggers)"
date: "July 9, 2016"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Data Science Steps

#### 1. Hypothesis Generation
Understanding the problem better by brainstorming possible factors that can impact the outcome

#### 2. Data Exploration
Looking at categorical and continuous feature summaries and making inferences about the data.

#### 3. Data Cleaning
Either filling in (with averages or default values) or removing missing values in the data and checking for outliers

#### 4. Feature Engineering
Modifying existing variables and creating new ones for analysis

#### 5. Model Building
Designing, Training, Testing / Running predictive models on the Data

#### 6. Visualizing & Communicating
Telling Stories from Data with summary and learned insights

While the steps 4 - 6 are more "tech-focused, jazzy/sexy and thus more appealing" to the younger minds, but the the steps 1 - 3 are often the key differentiators in the DataHack / Data Analytics Competitions and in Real Life... :-) ... 

And for the practical Business / Scientific Decision Making from Data Analytics projects, steps 1 to 3 usually takes 70% or more efforts ... :-) ...

I personally prefer the terms "Data Exploration" and "Data Cleaning" rather than "Data Manipulation & Exploration", which many Data Scientists casually use. Though I know many scientists often use the term "manipulation" in its true technical sense...

The connotation of the term "manipulation" especially "scares" me, as Data Scientists can really "twist / manipulate facts", if they lack "Conscience / Scientific Honesty"...!

For our R workshop, I request you to send the following R book (taken from a Coursera course, which most of us in the CC took) to all the registered participants:
I have already forwarded you the "Statistics in Plain English" book in another email ... :-)

[Coursera R Programming Course](https://www.coursera.org/learn/r-programming)
[Coursera R Programming Book](https://www.cs.upc.edu/~robert/teaching/estadistica/rprogramming.pdf)

I propose to hold two R workshops (each having 2-days, if possible), before Badrul Sarwar comes. 

The first R workshop should focus on the basics of Statistics, hands-on exercises using R, lab assignments, home-works, one micro DataHack challenge using the above 6 steps. 

The second R workshop should focus on reviewing the learning of the 1st workshop, moderate Statistics, Machine Learning in R, one mini DataHack challenge using the above 6 steps. 


I found the following URLs / PDFs / Books interesting: (Plz have a look...)

[Booz Allen Data Science Guide](https://www.boozallen.com/content/dam/boozallen/documents/2015/12/2015-FIeld-Guide-To-Data-Science.pdf)

[Applied Data Science Book of Columbia Uni](https://columbia-applied-data-science.github.io/appdatasci.pdf)

(light reading, easier approach, not too advanced...) [The Syracuse Uni Data Science Book](https://ischool.syr.edu/media/documents/2012/3/DataScienceBook1_1.pdf)

(and a lot more resources in the famous Kaggle Wiki that one can reasonably handle ...;-)
[A good list of Data Science Resources](https://www.kaggle.com/wiki/DataScienceBooksAndCourses)


## Inputs and Outputs

We can embed Shiny inputs and outputs in our R Markdown document, besides the stadalone ShinyApp (either merged in one `app.R` or in `server.R` and `ui.R`. 
Outputs are automatically updated whenever inputs change.  
This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. (which usually stays in the `server.R` file )
The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot. (which usually stays in the `ui.R` file )

```{r eruptions, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r tabsets, echo=FALSE}
shinyAppDir(
  system.file("examples/06_tabsets", package = "shiny"),
  options = list(
    width = "100%", height = 550
  )
)
```

Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.

### Violin Plots of the cars dataset
```{r cars, echo=FALSE}
library(vioplot)
x1 <- mtcars$mpg[mtcars$cyl==4]
x2 <- mtcars$mpg[mtcars$cyl==6]
x3 <- mtcars$mpg[mtcars$cyl==8]
vioplot(x1, x2, x3,
    names=c("4 cyl", "6 cyl", "8 cyl"),
    col="gold")
    title("Violin Plots of Miles Per Gallon")
```


