---
title: "Analysis of California Schools Test Data"
author: "Zaman (adopted from R-bloggers)"
date: "July 9, 2016"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Data Science Steps

#### 1. Hypothesis Generation
Understanding the problem better by brainstorming possible factors that can impact the outcome

#### 2. Data Exploration
Looking at categorical and continuous feature summaries and making inferences about the data.

#### 3. Data Cleaning
Either filling in (with averages or default values) or removing missing values in the data and checking for outliers

#### 4. Feature Engineering
Modifying existing variables and creating new ones for analysis

#### 5. Model Building
Designing, Training, Testing / Running predictive models on the Data

#### 6. Visualizing & Communicating
Telling Stories from Data with summary and learned insights

While the steps 4 - 6 are more "tech-focused, jazzy/sexy and thus more appealing" to the younger minds, but the the steps 1 - 3 are often the key differentiators in the DataHack / Data Analytics Competitions and in Real Life... :-) ... 

And for the practical Business / Scientific Decision Making from Data Analytics projects, steps 1 to 3 usually takes 70% or more efforts ... :-) ...

I personally prefer the terms "Data Exploration" and "Data Cleaning" rather than "Data Manipulation & Exploration", which many Data Scientists casually use. Though I know many scientists often use the term "manipulation" in its true technical sense...

The connotation of the term "manipulation" especially "scares" me, as Data Scientists can really "twist / manipulate facts", if they lack "Conscience / Scientific Honesty"...!

For our R workshop, I request you to send the following R book (taken from a Coursera course, which most of us in the CC took) to all the registered participants:
I have already forwarded you the "Statistics in Plain English" book in another email ... :-)

[Coursera R Programming Course](https://www.coursera.org/learn/r-programming)
[Coursera R Programming Book](https://www.cs.upc.edu/~robert/teaching/estadistica/rprogramming.pdf)

I propose to hold two R workshops (each having 2-days, if possible), before Badrul Sarwar comes. 

The first R workshop should focus on the basics of Statistics, hands-on exercises using R, lab assignments, home-works, one micro DataHack challenge using the above 6 steps. 

The second R workshop should focus on reviewing the learning of the 1st workshop, moderate Statistics, Machine Learning in R, one mini DataHack challenge using the above 6 steps. 


I found the following URLs / PDFs / Books interesting: (Plz have a look...)

[Booz Allen Data Science Guide](https://www.boozallen.com/content/dam/boozallen/documents/2015/12/2015-FIeld-Guide-To-Data-Science.pdf)

[Applied Data Science Book of Columbia Uni](https://columbia-applied-data-science.github.io/appdatasci.pdf)

(light reading, easier approach, not too advanced...) [The Syracuse Uni Data Science Book](https://ischool.syr.edu/media/documents/2012/3/DataScienceBook1_1.pdf)

(and a lot more resources in the famous Kaggle Wiki that one can reasonably handle ...;-)
[A good list of Data Science Resources](https://www.kaggle.com/wiki/DataScienceBooksAndCourses)


## Inputs and Outputs

We can embed Shiny inputs and outputs in our R Markdown document, besides the stadalone ShinyApp (either merged in one `app.R` or in `server.R` and `ui.R`. 
Outputs are automatically updated whenever inputs change.  
This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. (which usually stays in the `server.R` file )
The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot. (which usually stays in the `ui.R` file )

```{r eruptions, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## California School Data Analysis

```{r load_packages, echo=FALSE, message=FALSE, results='hide'}
# if necessary uncomment and install packages.
# install.packages("AER")
# install.packages("psych")
# install.packages("Hmisc")
# install.packages("ggplot2")
# install.packages("relaimpo")
library(AER) # interesting datasets
library(psych) # describe and psych.panels 
library(Hmisc) # describe
library(ggplot2) # plots: ggplot and qplot
library(relaimpo) # relative importance in regression
```

```{r load_data, echo=FALSE, message=FALSE, results='hide'}
# load the California Schools Dataset and give the dataset a shorter name
data(CASchools)
cas <- CASchools

# Convert grade to numeric

# table(cas$grades)
cas$gradesN <- cas$grades == "KK-08"

# Get the set of numeric variables
v <- setdiff(names(cas), c("district", "school", 
    			"county", "grades"))

```

### Q 1 What does the CASchools dataset involve?
Quoting the help (i.e., `?CASchools`), the data is "from all 420 K-6 and K-8 districts in California with data available for 1998 and 1999" and the variables are:

    * district: character. District code.
    * school: character. School name.
    * county: factor indicating county.
    * grades: factor indicating grade span of district.
    * students: Total enrollment.
    * teachers: Number of teachers.
    * calworks: Percent qualifying for CalWorks (income assistance).
    * lunch: Percent qualifying for reduced-price lunch.
    * computer: Number of computers.
    * expenditure: Expenditure per student.
    * income: District average income (in USD 1,000).
    * english: Percent of English learners.
    * read: Average reading score.
    * math: Average math score.

Let's look at the basic structure of the data frame. i.e., the number of observations and the types of values:

```{r echo=FALSE, message=FALSE, results='hide'}
str(cas)
# Hmisc::describe(cas) # For more extensive summary statistics
```



### Q. 2  To what extent does expenditure per student vary?
```{r cas2, echo=FALSE, message=FALSE, results='hide'}
qplot(expenditure, data = cas) + xlim(0, 8000) + 
		xlab("Money spent per student ($)") +
		ylab("Count of schools")

round(t(psych::describe(cas$expenditure)), 1)
```

The greatest expenditure per student is around double that of the least expenditure  per student.


### Q. 3a  What predicts expenditure per student?
```{r echo=FALSE, message=FALSE, results='hide'}
# Compute and format set of correlations
corExp <- cor(cas["expenditure"], 
		cas[setdiff(v, "expenditure")])
corExp <- round(t(corExp),2)
corExp[order(corExp[,1], decreasing = TRUE), , 
		drop = FALSE]
```

More is spent per student in schools :

1. where people with greater incomes live
2. reading scores are higher
3. that are K-6


### Q. 4  what is the relationship between district level maths and reading scores?
```{r cas4, echo=FALSE, message=FALSE, results='hide'}
ggplot(cas, aes(read, math)) + geom_point() + 
		geom_smooth()
```

At the district level, the correlation is very strong (r = The correlation is `r round(cor(cas$read, cas$math), 2)`). From prior experience I'd expect correlations at the individual-level in the .3 to .6 range.  Thus, these results are consistent with group-level relationships  being much larger than individual-level relationships.

### Q. 5 What is the relationship between maths and reading after partialling out other effects?


```{r echo=FALSE, message=FALSE, results='hide'}
# command has strange syntax requiring column numbers rather than variable names
partial.r(cas[v], 
          c(which(names(cas[v]) == "read"), which(names(cas[v]) == "math")), 
          which(!names(cas[v]) %in% c("read", "math"))
          )
```

The partial correlation is still very strong but is substantially reduced.


### Q. 6 What fraction of a computer does each student have?
```{r echo=FALSE, message=FALSE, results='hide'}
cas$compstud <- cas$computer / cas$students
describe(cas$compstud)
qplot(compstud, data = cas)
```

The mean number of computers per student is `r round(mean(cas$compstud), 3)`.


### Q. 7 What is a good model of the combined effect of other variables on academic performance (i.e., math and read)?
```{r cas7, echo=FALSE, message=FALSE, results='hide'}
# Examine correlations between variables
psych::pairs.panels(cas[v])
```

`pairs.panels` shows correlations in the upper triangle, scatterplots in the lower triangle, and variable names and distributions on the main diagonal.
After examining the plot several ideas emerge.

```{r cas7.transformation, echo=FALSE, message=FALSE, results='hide', tidy=FALSE}
# (a) students is a count and could be log transformed
cas$studentsLog <- log(cas$students)

# (b) teachers is not the variable of interest:
#	it is the number of students per teacher
cas$studteach <- cas$students /cas$teachers
# (c) computers is not the variable of interest:
#  it is the ratio of computers to students
# table(cas$computer==0) 
# Note some schools have no computers so ratio would be problematic.
# Take percentage of a computer instead
cas$compstud <- cas$computer / cas$students 

# (d) math and reading are correlated highly, reduce to one variable
cas$performance <- as.numeric(
		scale(scale(cas$read) + scale(cas$math)))
```
Normally, I'd add all these transformations to an initial data transformation file that I call in the first block, but for the sake of the narrative, I'll leave them here.

Let's examine correlations between predictors and outcome.
```{r echo=FALSE, message=FALSE, results='hide'}
m1cor <- cor(cas$performance, 
		cas[c("studentsLog", "studteach",	"calworks",  
						"lunch", "compstud", "income", 
						"expenditure", "gradesN")])
t(round(m1cor, 2))
```


Let's examine the multiple regression.
```{r echo=FALSE, message=FALSE, results='hide'}
m1 <- lm(performance ~ studentsLog + studteach + 
				calworks + lunch + compstud
				+ income + expenditure + grades, data = cas)		
summary(m1)
```
And some indicators of predictor relative importance.
```{r echo=FALSE, message=FALSE, results='hide'}
# calc.relimp from relaimpo package.
(m1relaimpo <- calc.relimp(m1,	type="lmg",	rela=TRUE))
```

Thus, we can conclude that:

1. Income and indicators of income (e.g., low levels of lunch vouchers) are the two main predictors. Thus, schools with greater average income tend to have better student performance.
2. Schools with more computers per student have better student performance.
3. Schools with fewer students per teacher have better student performance.

For more information about relative importance and the `relaimpo` package measures check out [Ulrike Grömping's website](http://prof.beuth-hochschule.de/groemping/relaimpo/).
Of course this is all observational data with the usual caveats regarding causal interpretation.

## Now, let's look at some weird stuff.
### Q. 8.1 What are common words in Californian School names?

```{r echo=FALSE, message=FALSE, results='hide'}
# create a vector of the words that occur in school names
lw <- unlist(strsplit(cas$school, split = " "))

# create a table of the frequency of school names
tlw <- table(lw)

# extract cells of table with count greater than 3
tlw2 <- tlw[tlw > 3]

# sorted in decreasing order
tlw2 <- sort(tlw2, decreasing = TRUE)

# values as proporitions
tlw2p <- round(tlw2 / nrow(cas), 3)

# show this in a bar graph
tlw2pdf <- data.frame(word = names(tlw2p), 
		prop = as.numeric(tlw2p),
		stringsAsFactors = FALSE)

## Error: stat_count() must not be used with a y aesthetic.
### the deafult arg in geom_bar() is stat=stat_count() ...

## ggplot(tlw2pdf, aes(word, prop)) + geom_bar() + coord_flip()

ggplot(tlw2pdf, aes(word, prop)) + geom_bar(stat="identity") + coord_flip()
```

```{r echo=FALSE, message=FALSE, results='hide'}
# make it log counts
# stat="identity"
# ggplot(tlw2pdf, aes(word, log(prop*nrow(cas)))) + 
# 		geom_bar() + coord_flip()

ggplot(tlw2pdf, aes(word, log(prop*nrow(cas)))) + 
		geom_bar(stat="identity") + coord_flip()
```

The word "Elementary" appears in almost all school names (`r round(100 * tlw2p["Elementary"], 1)`%).  The word "Union" appears in around half (`r round(100 * tlw2p["Union"], 1)`%).

Other common words pertain to:

* Directions (e.g., South, West), 
* Features of the environment 
    (e.g., Creek, Vista, View, Valley)
* Spanish words (e.g., rio for river; san for saint)


### Q. 8.2 Is the number of letters in the school's name related to academic performance?
```{r echo=FALSE, message=FALSE, results='hide'}
cas$namelen <- nchar(cas$school)
table(cas$namelen)
round(cor(cas$namelen, cas[,c("read", "math")]), 2)
```
The answer appears to be "no".


### Q.  8.3 Is the number of words in the school name related to academic performance?
```{r echo=FALSE, message=FALSE, results='hide'}
cas$nameWordCount <- 
		sapply(strsplit(cas$school, " "), length)
table(cas$nameWordCount)
round(cor(cas$nameWordCount, cas[,c("read", "math")]), 2)
```
The answer appears to be "no".


### Q. 8.4 Are schools with nice popular nature words in their name doing better academically?
```{r echo=FALSE, message=FALSE, results='hide'}
tlw2p #recall the list of popular names
```

```{r echo=FALSE, message=FALSE, results='hide'}
# Create a quick and dirty list of popular nature names
naturenames <- c("Valley", "View", "Creek", 
		"Lake", "Mountain",	"Park", "Rio", 
		"Vista", "Grove", "Lakeside")

# work out whether the word is in the school name
schsplit <- strsplit(cas$school, " ")
cas$hasNature <- sapply(schsplit, 
		function(X) length(intersect(X, naturenames)) > 0) 
round(cor(cas$hasNature, cas[,c("read", "math")]), 2)
```
So we've found a small correlation.  
Let's graph the data to see what it means:

```{r echo=FALSE, message=FALSE, results='hide'}
ggplot(cas, aes(hasNature, read)) + 
        geom_boxplot() + 
		geom_jitter(position=position_jitter(width=0.1)) +
		xlab("Has a nature name") +
		ylab("Mean student reading score")
```
So in the sample nature schools have slightly better reading score (and if we were to graph it, maths scores). However, the number of schools having nature names is actually somewhat small (n= `r sum(cas$hasNature)`) despite the overall quite large sample size.

But is it statistically significant?
```{r echo=FALSE, message=FALSE, results='hide'}
t.read <- t.test(cas[cas$hasNature, "read"], cas[!cas$hasNature, "read"])
t.math <- t.test(cas[cas$hasNature, "math"], cas[!cas$hasNature, "math"])
```
So, the p-value is less than .05 for reading (p = `r round(t.read$p.value, 3)`) but not quite for maths (p = `r round(t.math$p.value, 3)`).  Bingo!  After a little bit of data fishing we have found that reading scores are "significantly" greater for those schools with the listed nature names.

**But wait**: I've asked three separate exploratory questions or perhaps six if we take maths into account.

* $\frac{.05}{3} =$ `r 0.05 / 3`
* $\frac{.05}{6} =$ `r 0.05 / 6`

At these Bonferonni corrected p-values,  the result is non-significant. Oh well...


## Review
Anyway, the aim of this post was not to make profound statements about California schools. Rather the aim was to show how easy it is to produce quick reproducible reports with R Markdown. If you haven't already, you may want to open up the R Markdown file used to produce this post in RStudio, and compile the report yourself.

In particular, I can see R Markdown being my tool of choice for:

* Blog posts
* Posts to StackExchange sites
* Materials for training workshops
* Short consulting reports, and
* Exploratory analyses as part of a larger project.

The real question is how far I can push Markdown before I start to miss the control of LaTeX.  Markdown does permit arbitrary HTML. Anyway, if you have any thoughts about the scope of R Markdown, feel free to add a comment.

### Violin Plots of the cars dataset
```{r violinplot, echo=FALSE, message=FALSE, results='hide'}
library(vioplot)
x1 <- mtcars$mpg[mtcars$cyl==4]
x2 <- mtcars$mpg[mtcars$cyl==6]
x3 <- mtcars$mpg[mtcars$cyl==8]
vioplot(x1, x2, x3,
    names=c("4 cyl", "6 cyl", "8 cyl"),
    col="gold")
    title("Violin Plots of Miles Per Gallon")
```


